---
title: "Power for mediation"
author: Marcello Gallucci 
topic: mediation
category: help
bibliography: 
     - bib.bib
link-citations: true
        
---


```{r echo=FALSE,results='hide'}
library(mcdocs)
mcdocs_init()

```

`r version("0.4.0")`

`r pic("pics/mediation/ui.png")`

`Mediation power analysis` allows for computing power parameters for testing indirect (mediated) effects in several mediational models. It can be used to estimate power parameters based on the `Sobel test`, `joint significance` or `Monte Carlo` method. 

`r incomplete`


The initial step is to determine the objective of the analysis ( _aim_ ). Further details regarding different objectives can be found in the `Info` results field.

`r pic("pics/common/aim.1.png")`


The second step is to decide what type of model we are dealing with. 

`r pic("pics/mediation/input.1.png")`


# Models

* __Simple mediation (3      variables)__ : This is the classical simple mediation, with one independent variable, one mediator. The model looks like this
`r pic("pics/mediation/diagram.1.png")`

* __Complex (multiple mediators)__ : Allows computation of power parameters for more complex models: two parallel mediator, three parallel mediator and serial mediation with two mediators. They are, respectively:

`r pic("pics/mediation/diagram.2.png")`



# Simple Mediation Model

In simple mediation models there is one independent variable, one mediator and one dependent variable. The model is:

`r pic("pics/mediation/diagram.1.png")`

The `a` coefficient is the simple standardized coefficient of a regression from X to M (mediator). `b` and `c'` are the standardized coefficients of a multiple regression with $Y$ as dependent variable and $X$ and $M$ as independent variables. The mediated effect $ME=a*b$ is given by the product of the coefficients along the mediator pathway. 

As for any `r modulename()` command, we can alway obtain some help in setting the analysis by clicking on the `plus` icon in the `Info` bar.

`r pic("pics/mediation/output.0.1.png")`
`r pic("pics/mediation/output.0.2.png")`


Thus, we now declare what are the expected standardized coefficients of the model.

`r pic("pics/mediation/input.2.png")`

The power `r opt_label("Parameters")` are the usual parameters of power analysis.


```{r child60, child='commons/pparameters_chunk.Rmd'}
```

# Main results

As soon as the coefficients are set, the path diagram is updated and we obtain the results in the `A Priori Power Analysis` table.

`r pic("pics/mediation/output.1.png")`

The table recaps the input parameters plus the estimated parameter we asked in the `r opt("Compute")` option. In this example, we asked for `N`, so the module estimated $N=223$ as the required sample size to obtain a power of .90 or more, given $a=.3$, $b=.3$ and $c'=0$. An additional textual description of the results can be obtained by selecting `r opt("Exploratory text")` under the `r opt_label("Options")` label.

`r pic("pics/mediation/output.text.png")`


# Tests

`r pic("pics/mediation/input.3.png")`


## Joint significance

By default, `r modulename()` assumes that the `r opt("joint significance")` method [@yzerbyt2018new] will be used to test the mediated effect. This method determines the significance of the mediated effect by testing whether both paths $a$ and $b$ are significant: if both are significant, the mediated effect is considered significant. Additionally, this method produces results very similar to the Monte Carlo method [@yzerbyt2018new], while being much faster in terms of computation.

In our example, by choosing `r opt("joint significance")`, we obtain a required sample size of $N=137$.

`r pic("pics/mediation/output.1.png")`

## Sobel test

Alternatively, users can opt for the Sobel test [@sobel1982asymptotic] to evaluate the significance of the mediated effect. The Sobel test is commonly used for mediation analysis due to its computational efficiency, as it is based on the z-test. However, it is important to note that while the Sobel test is fast, it is known to introduce some bias and may not always be the most reliable method for testing mediated effects. In our example, we obtain an N=223, which is larger than the required N obtained with the `joint significance` test, testifying the lower power of the Sobel test [@mackinnon2002comparison].

`r pic("pics/mediation/output.2.png")`

## Monte Carlo

The third method is the `r opt("Monte Carlo")` method. This approach is based on simulations [@mackinnon2004confidence]. It generates a distribution of parameters with the appropriate standard error and evaluates their joint probability of being significant at the error rate provided in the input. This simulation process is repeated $R$ times, and the probability of obtaining a significant result is calculated from the R simulations. If one intends to assess significance using a bootstrap method [@preacher2008asymptotic], the Monte Carlo method is an accurate approximation of power parameters [@zhang2014monte].

For the `r opt("Monte Carlo")` method, the number of simulations ($R$) can be set in the `r opt_panel("Options")`.  When the number of simulations exceeds 1000, it is recommended to enable `r opt("Parallel computation")` to speed up the process. However, users should be aware that in certain operating systems (such as Windows), parallel computation might not function properly.

`r pic("pics/mediation/input.4.png")`

The Monte Carlo (MC) and joint significance methods yield nearly identical required sample sizes (N) in our example. While MC results may slightly vary between runs, this variability decreases as the number of simulations (R) increases, improving estimate precision. Typically, both methods produce very similar results.

`r pic("pics/mediation/output.3.png")`

# Sensitivity Analysis

Sensitivity analysis, exploring different plots of possible combinations of parameters, can be carried out like for any other `r modulename()` sub-modules. Please visit `r link_pages(nickname="sensitivity")` page for more details.



`r backto()`

`r issues()`
