---
title: "Independent sample T-test: Equivalence tests"
topic: ttest
category: rosetta
nickname: ttest_ind_equi
output: 
  html_document:
     includes:
         in_header: ganalytics.txt
     toc: true
     toc_float:
        collapsed: false
bibliography: 
     - bib.bib
link-citations: true
        
---


```{r echo=FALSE,results='hide'}
library(mcdocs)
mcdocs_init()

```

# Power Analysis

`r version("0.2.0")` 

Here we compare the results of `r modulename()` with published results and other software that performs equivalence tests for the t-test.

# Independent sample t-test

For the independent t-test, we will refer to @lakens2017equivalence `r ext_url("Table 1","https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5502906/table/table1-1948550617697177/?report=objectonly")`
, in which sample size are reported for an equivalence test with expected value $\delta=0$ and varying "Bound", in `r modulename()` refereed to as 'equivalence limit'. @lakens2017equivalence Table 1 reports results for $\alpha=.05$ and $\alpha=.01$.  The table (available `r ext_url("here","https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5502906/table/table1-1948550617697177/?report=objectonly")`) is reproduced below.

`r pic("rosetta/pics/ttest_ind_equi_1.png")`

The table reports the required N (per group) to obtain a sufficient power (80% or 90%) given a certain $\alpha$ and a certain limit $\Delta$ of equivalence.

To compare the results with R, we are going to use results from `r ext_url("powerandsamplesize.com website", "http://powerandsamplesize.com/Calculators/Compare-2-Means/2-Sample-Equivalence")`, in which the R code of their nice online calculator is public and open, so we could implemented in R to make comparisons.

### PAMLj

For the firs test, we insert an `r opt("Equivalence limit")` of $.1$, an expected $\delta=0$, power=.80 and $\alpha=.05$.
`r pic("rosetta/pics/ttest_ind_equi_2.png")`
`r pic("rosetta/pics/ttest_ind_equi_3.png")`

As expected, the results are quite similar. The 1-unit difference observed (1714 vs 1713) is simply due to the rounding, which may yield slightly different results across different software. 

### R

```{r }

muA=5
muB=5
delta=1
kappa=1
sd=10
alpha=0.05
beta=0.20
nB=(1+1/kappa)*(sd*(qnorm(1-alpha)+qnorm(1-beta/2))/(abs(muA-muB)-delta))^2
z=(abs(muA-muB)-delta)/(sd*sqrt((1+1/kappa)/nB))
ceiling(nB) # 108
(Power=2*(pnorm(z-qnorm(1-alpha))+pnorm(-z-qnorm(1-alpha)))-1)
```

For $\Delta=.2$ we proceed as before.

### PAMLj

For the firs test, we insert an <b>Equivalence limit</b> of $.1$, an expected $\delta=0$, power=.80 and $\alpha=.05$.
`r pic("rosetta/pics/ttest_ind_equi_4.png")`
`r pic("rosetta/pics/ttest_ind_equi_5.png")`

### R

```{r }

muA=5
muB=5
delta=2
kappa=1
sd=10
alpha=0.05
beta=0.20
nB=(1+1/kappa)*(sd*(qnorm(1-alpha)+qnorm(1-beta/2))/(abs(muA-muB)-delta))^2
z=(abs(muA-muB)-delta)/(sd*sqrt((1+1/kappa)/nB))
ceiling(nB) # 108
(Power=2*(pnorm(z-qnorm(1-alpha))+pnorm(-z-qnorm(1-alpha)))-1)
```

Again, results are identical even after rounding.

As a final test, we try $\Delta=.5$, $\alpha=.01$, power=.90, expecting  $N=128$ (see table under "exact" columns).

### PAMLj

For the firs test, we insert an <b>Equivalence limit</b> of $.1$, an expected $\delta=0$, power=.80 and $\alpha=.05$.
`r pic("rosetta/pics/ttest_ind_equi_6.png")`
`r pic("rosetta/pics/ttest_ind_equi_7.png")`

```{r }

muA=5
muB=5
delta=5
kappa=1
sd=10
alpha=0.01
beta=0.10
nB=(1+1/kappa)*(sd*(qnorm(1-alpha)+qnorm(1-beta/2))/(abs(muA-muB)-delta))^2
z=(abs(muA-muB)-delta)/(sd*sqrt((1+1/kappa)/nB))
ceiling(nB) # 108
(Power=2*(pnorm(z-qnorm(1-alpha))+pnorm(-z-qnorm(1-alpha)))-1)
```


# Navigate

Back to `r link_pages(nickname="rosetta")`


`r issues()`

# References
'