---
title: "GLM: posthoc power"
topic: posthoc
category: rosetta
nickname: ros_glm_posthoc
linklabel: "Rosetta: GLM posthoc power" 
bibliography: 
     - bib.bib
link-citations: true
        
---


```{r echo=FALSE,results='hide'}
library(mcdocs)
mcdocs_init()
```

# Power Analysis: post hoc power 

`r version("0.1.0")` 

Here we compare the results of `r modulename()` with other software that performs power analysis. In particular, we will compare our results with [R pwr package](https://cran.r-project.org/web/packages/pwr/pwr.pdf) and [G*Power](https://www.psychologie.hhu.de/arbeitsgruppen/allgemeine-psychologie-und-arbeitspsychologie/gpower). 

We use a few examples from [G*Power manual](https://www.psychologie.hhu.de/arbeitsgruppen/allgemeine-psychologie-und-arbeitspsychologie/gpower), then we test some other design and analysis. In all these example the aim of the analysis is to find the post hoc power given an effect size, a sample size and the design characteristics.

# Multiple regression
<i> Multiple Regression - omnibus (deviation of R 2 from zero), fixed model, p. 33 </i>

Here we estimate the post-hoc power for the F-test associated to the $R^2$ of a model with 5 predictors. Five predictors means that the model degrees of freedom are 5. Because the test is on the whole model $R^2$, also the the effect DF are 5. The $R^2$ is transformed into the $f^2$ effect size with the simple formula $f^2=R^2/(1-R^2)$.

*   Aim = power
*   Expected $R^2$    = .10
*   $f^2$              = `r round(0.10/(1-.10),digits=4)`
*   Sample Size (N) = 95
*   Effect DF       = 5
*   Model  DF       = 5
*   Alpha           = .05

### G*Power

`r pic("rosetta/pics/gpower.glm.1.png")`

    
### R

In R, we can use the `pwr.f2.test` command, just plugging in the effect size $f^2$ and the effect DF $u=5$. However, $N$ should be inserted as the residuals degrees of freedom, thus we have $v=N-u-1=95-5-1=89$

```{r }

pwr::pwr.f2.test(f2=.1111,u=5, v=89,sig.level=.05)

```

### PAMLj

In `r modulename()` we can use both `r opt("Partial Eta-Squared")` or `r opt("Eta-Squared")`, because in the case of the $R^2$ the two effect size are identical. For simplicity, we use `r opt("Partial Eta-Squared")`

`r pic("rosetta/pics/glm.1.png")`

Had we used `r opt("Eta-Squared")`, we simply put the $R^2$ to test both in the effect size and the $R^2$ field, obtaining the same results.

`r pic("rosetta/pics/glm.1.2.png")`

We obtain the same results in packages, although we will see that this is not always the case (see next example).


# Increase in $R^2$

<i> F test: Multiple Regression - special (increase of R2), fixed model , p. 37 </i>

The first example examines a linear model (regression) with $R_f^2=.30$ with 9 predictors and compare it to a nested model with $R_n^2=.25$ with 5 predictors. Basically, the test is aimed at testing the variance explained by 4 predictors present in the full model but not the nested model. The variance explained by the 4 predictors is therefore $R_f^2-R_n^2=.30-.25=.05$.  This is also called $\Delta R^2$, but more formally is a $\eta^2$ with 4 degrees of freedom evaluated in a model with 9 degrees of freedom.

The $\eta^2$ is transformed into the $f^2$ effect size with the simple formula $f^2=(R_f^2-R_n^2)/(1-R_f^2)$.

*   Aim = power
*   $R_f^2$    = .30
*   $R_n^2$    = .25
*   $\eta^2$    =  0.05
*   $f^2$              = `r round(0.05/(1-.30),digits=5)`
*   Sample Size (N) = 90
*   Effect DF       = 4
*   Model  DF       = 9
*   Alpha           = .01

### G*Power

`r pic("rosetta/pics/gpower.glm.2.png")`

### R

In R, we can use the `pwr.f2.test` command, just plugging in the effect size $f^2$ and the effect DF $u=4$. Keeping in mind that the model has $m=9$ degrees of freedom, the $N$ is inserted as the residuals degrees of freedom, that is $v=N-m-1=90-9-1=80$

```{r }
f2=.05/(1-.30)
pwr::pwr.f2.test(f2=f2,u=4, v=80,sig.level=.01)

```

Notice that the results (_power_) are similar, but not as close as before. This is because and G*Power and `pwr:pwr.f2.test()` use two different computations of the F-distribution _non centrality parameter $\lambda"_. G*Power uses $\lambda=f^2 \cdot N$, whereas  `pwr:pwr.f2.test()` uses  $\lambda=f^2 \cdot (u+v+1)$.  `r modulename()` allows to choose which one to use. To obtain the correct power one should adjust the effect size such that the non-centrality parameter will be correct. One can do this:

```{r }
N<-90
u<-4
v=80
f2=.05/(1-.30)
f2adj<-f2*N/(v+u+1)
pwr::pwr.f2.test(f2=f2adj,u=u, v=v,sig.level=.01)

```

The power is now identical to the one estimated in GPower.

### PAMLj

By default,  `r modulename()` employs the GPower _non centrality parameter_, yielding the same results as GPower.

`r pic("rosetta/pics/glm.2.1.png")`

BY de-selecting the option `r opt("G*Power NCP")`, we obtain the same results as in R without adjustment.

`r pic("rosetta/pics/glm.2.2.png")`


# ANOVA 

<i> F test: Fixed effects ANOVA - special, main effects and interaction p. 27</i>

Here we have an ANOVA design $A \times B \times C$ with $A$ and $B$ having 3 levels and $C$ with 4 levels. Thus the overall model features 36 groups, with $A$ and $B$ effects having 2 degrees of freedom, and $C$ having 3 DF. The researcher is interested in the power of the test associated with the interaction $A*B$, with a partial Eta-squared $p\eta^2=0.05665751$ (here the value is slightly different than in GPower manual, but this value honors the f value used in the manual). The interaction will have $DF=2*2=4$ degrees of freedom, and the whole model will have $DF=35$, the number of groups minus one. Here are the parameters.

*   Aim = power
*   $p\eta^2$   = 0.05665751
*   $f$              = `r round(sqrt(0.05665751/(1-0.05665751)),digits=5)`
*   Sample Size (N) = 90
*   Effect DF       = 4
*   Model  DF       = 35
*   Alpha           = .05

### G*Power

First notice that for this problem, G*Power offers  _F test: Fixed effects ANOVA_ routine, which requires an $f$ as the effect size . $f$ is simply the square root of $f^2$ so it can be computed from the partial Eta-squared as  $f=\sqrt{p\eta^2/(1-p\eta^2)}$ .

Plugging in the parameters we obtain:

`r pic("rosetta/pics/gpower.glm.3.png")`



### R

As we have seen before, `pwr.f2.test` command will underestimate the power due to the fact that the non centrality parameter is smaller than the one used in G*Power.  Nonetheless, we can get an estimation of the power by adjusting the effect size $f^2$ such as to match the correct NCP will be $\lambda=Nf^2$. The effect DF $u=4$ and the model DF are $m=35$, yielding $v=N-m-1=108-35-1=72$

```{r }
f<-0.2450722
u<-4
m<-35
N=108
v<-N-m-1
f2<-f^2*N/(u+v+1)
pwr::pwr.f2.test(f2=f2,u=4, v=72,sig.level=.05)

```

We obtained the same results as in G*Power

### PAMLj

We plug in the partial Eta-squared and the required parameters, and we obtain the same results as before.

`r pic("rosetta/pics/glm.3.png")`



`r issues()`

# Back to `r link_pages(nickname="rosetta")`


'