---
title: "Mediation: Complex models"
topic: mediation
category: rosetta
nickname: ros_mediation_complex
bibliography: 
     - bib.bib
link-citations: true
        
---


```{r echo=FALSE,results='hide'}
library(mcdocs)
mcdocs_init()
```


:::{ .version }
0.4.0
:::


Here we compare the results of `r modulename()` with other software that performs power analysis for mediation in complex models. We consider three complex models: two parallel mediators, three parallel mediators, and two sequential mediators. At the moment, the only package that explicitly deals with these models is the [Monte Carlo Power Analysis](https://schoemanna.shinyapps.io/mc_power_med/) Shiny app, which provides power estimates based solely on simulations. However, with a bit of work, one can also employ the [pwrss R package](https://CRAN.R-project.org/package=pwrss) to test parameters for these models. The advantage of `pwrss` is that it provides estimates based on the Sobel test, joint significance, and Monte Carlo methods, allowing for better comparisons of the results.

* [pwrss R package](https://CRAN.R-project.org/package=pwrss) provides desired power and required sample size based on Sobel test and joint significance methods. It also provides desired power based on simulations. The simulations are based on Monte Carlo parametric re-sampling.

* [Monte Carlo Power Analysis](https://schoemanna.shinyapps.io/mc_power_med/) shiny app provides desired power based on simulations. The simulations are based on Monte Carlo semi-parametric re-sampling.

We test the software finding power, because the two packages readily estimate the parameter.

# Two parallel mediators

`r pic("pics/mediation/complex.input.1.png")`

`r pic("pics/mediation/complex.diagram.1.png")`

:::{ .adm .adm-setup}
Setup

*   Aim = Power
*   Expected $a1$    = .3
*   Expected $b1$    = .3
*   Expected $a2$    = .4
*   Expected $b2$    = .2
*   Expected $r12$   = .25  __expected correlations among mediators__
*   Expected $c'$    = .15
*   N                = 100
*   Alpha            = .05

:::

`r pic("rosetta/pics/mediation/complex.input.1.png")`

For `Joint significance` method:

`r pic("rosetta/pics/mediation/complex.output.1.png")`

For `Sobel test` method:

`r pic("rosetta/pics/mediation/complex.output.2.png")`

For `Monte Carlo` method:

`r pic("rosetta/pics/mediation/complex.output.3.png")`

As expected, `joint significance` and `Monte Carlo` methods agree in the estimated parameters, whereas `Sobel test` reports a lower power for both mediated effects as compared with the other methods estimation.


## Monte Carlo shiny app

By setting up the same parameters in the `Monte Carlo Shiny app`, we obtain very similar results, although the app seems to produce smaller power values than `r modulename()`. The reason for this small but consistent discrepancy is the way different software computes the expected standard errors of the coefficients. Both `r modulename()` and `pwrss` simulate a z-test when computing standard errors and the distribution of the parameters. The `Monte Carlo Shiny app`, however, assumes normality of the parameters but computes the standard errors based on the t-test. For small samples, this difference may yield slightly different final parameters.

`r pic("rosetta/pics/mediation/complex.mc.input.1.png")`
`r pic("rosetta/pics/mediation/complex.mc.output.1.png")`

## pwrss

To obtain results for the two parallel mediators model using `pwrss`, we need some preliminary computations. The `pwrss.z.mediation` function allows you to input, for each mediated effect, the expected R-squared in predicting the mediator (option `r2m.x` in the function) and the R-squared in predicting the dependent variable (option `r2y.mx` in the function). This is because the standard error of the coefficients depends on the R-squared values, and so does the standard error of the mediated effect. In other words, the presence of other mediators is taken into account in the computation by including their effects in the R-squared predicting the dependent variable. When the R-squared values are correctly specified, the pwrss.z.mediation function can be used to estimate power (or N) for complex models as well.

Let's start with the first mediated effect $ME=a_1 \cdot b_1$. The computation is easy: $R_{m.x}^2=a_1^2$, because $a_1$ is basically a Pearson correlation. For $R_{y.mx}^2$ (predicting Y), the computation is more complex. We can take the R-squared from `r modulename()` results.

`r pic("rosetta/pics/mediation/complex.output.4.png")`

The R-squared is $R_{y.mx}^2=.2335$. Plugging the parameters in we obtain the results, again in lines with the `r modulename()` results.

```{r}
a1  <- .3
b1  <- .3
a2  <- .4
b2  <- .2
r1  <- .25
cp <- .15
r2y.mx<-.2335

pwrssp<-pwrss::pwrss.z.mediation(a=a1,b=b1,r2y.mx=r2y.mx,n=100,alpha=.05)

```

For the second mediated effect, we have

```{r}
a1  <- .3
b1  <- .3
a2  <- .4
b2  <- .2
r1  <- .25
cp <- .15
r2y.mx<-.2335

pwrssp<-pwrss::pwrss.z.mediation(a=a2,b=b2,r2y.mx=r2y.mx,n=100,alpha=.05)

```

Nicely in line with `r modulename()` results.

Admittedly, taking the R-squared from `r modulename()` and plug it into another software to check the results of `r modulename()` it is not much of a test. So let's check the R-squared. 

In multiple regression, the $R^2$ can be computed as:
$$
R^2=\sum{\beta_i \cdot r_{yi}}
$$
Now, we have the __betas__, we need the correlations. For the mediators let's indicate as $1$ and $2$ the two mediators. 

For mediator $M1$ we have that

$$
 r_{YM1}=b_1+b_2 \cdot r_{12} + c' \cdot a_1
$$

For mediator $M2$ we have that


$$
 r_{YM2}=b_2+b_1 \cdot r_{12} + c' \cdot a_2
$$


For mediator $X$ we have that


$$
 r_{YX}=c'+a_1 \cdot b_1 + a_2 \cdot b_2  
$$
Substituting the correlations into the $R^2$ formula, we obtain

$$
R^2 = b_1^2 + b_2^2 + c^2 + 2 \cdot b_1 \cdot b_2 \cdot r_{12} + 2 \cdot a_1 \cdot b_1 \cdot c + 2 \cdot a_2 \cdot b_2 \cdot c
$$

Numerically:

```{r}
a1  <- .3
b1  <- .3
a2  <- .4
b2  <- .2
r12  <- .25
cp <- .15

rym1<-b1+b2*r12+cp*a1
rym2<-b2+b1*r12+cp*a2
ryx<- cp +a1*b1+a2*b2

b1^2+b2^2+cp^2+ 2*b1*b2*r12+ 2 * a1*b1*cp+2*a2*b2*cp


```

So, it's fine.

# Three parallel mediators

We now add another mediator

:::{ .adm .adm-setup}
Setup

*   Aim = Power
*   Expected $a1$    = .3
*   Expected $b1$    = .3
*   Expected $a2$    = .4
*   Expected $b2$    = .2
*   Expected $a3$    = .25
*   Expected $b3$    = .15
*   Expected $r12$   = .25 
*   Expected $r13$   = .1 
*   Expected $r23$   = .4 
*   Expected $c'$    = .15
*   N                = 100
*   Alpha            = .05

:::


`r pic("rosetta/pics/mediation/complex.diagram.2.png")`

## PAMLj

`r pic("rosetta/pics/mediation/complex.input.2.png")`

For `Joint significance` method:

`r pic("rosetta/pics/mediation/complex.output.5.png")`

For `Sobel test` method:

`r pic("rosetta/pics/mediation/complex.output.6.png")`

For `Monte Carlo` method:

`r pic("rosetta/pics/mediation/complex.output.7.png")`

## pwrss

To utilize `pwrss` we need to input, for each mediated effect, the $R^2$ predicting both the dependent variable and the mediators. We can take these estimates from `r modulename()`.

`r pic("rosetta/pics/mediation/complex.output.8.png")`

```{r}
a1  <- .3
b1  <- .3
a2  <- .4
b2  <- .2
a3  <- .25
b3  <- .15
cp <- .15

r2y.mx<-.3003
rmx<-a1^2
pwrssp<-pwrss::pwrss.z.mediation(a=a1,b=b1,r2y.mx=r2y.mx, r2m.x=rmx  ,n=100,alpha=.05)

rmx<-a2^2
pwrssp<-pwrss::pwrss.z.mediation(a=a2,b=b2,r2y.mx=r2y.mx, r2m.x=rmx  ,n=100,alpha=.05)

rmx<-a3^2
pwrssp<-pwrss::pwrss.z.mediation(a=a3,b=b3,r2y.mx=r2y.mx, r2m.x=rmx  ,n=100,alpha=.05)

```

## Monte Carlo shiny app

Unfortunately, to obtain the correct power parameters in the app, we need to input the correlation matrix. Obtaining the correlation matrix starting from the betas coefficients is tedious, so I just list it here.


```{r}
sigma<-matrix(c(1.0000, 0.30, 0.400, 0.2500, 0.3575
, 0.3000, 1.00, 0.250, 0.1000, 0.4100
, 0.4000, 0.25, 1.000, 0.4000, 0.3950
, 0.2500, 0.10, 0.400, 1.0000, 0.2975
, 0.3575, 0.41, 0.395, 0.2975, 1.0000),ncol=5,nrow=5)

sigma
```

To be sure that they are the correct correlations, just check if they produced the intended beta coefficients.

```{r}

data<-as.data.frame(MASS::mvrnorm(100,c(0,0,0,0,0),sigma, empirical=T))
names(data)<-c("x","m1","m2","m3","y")

zapsmall(coef(lm(y~m1+m2+m3+x,data=data)))

```
as expected, they are correct.


At this point, we can insert the correlation in the app.


`r pic("rosetta/pics/mediation/complex.mc.input.2.png")`

`r pic("rosetta/pics/mediation/complex.mc.output.2.png")`

As expected, the power obtained is slightly lower than the one obtained with `pwrss` and `r modulename()`, but it is substantially in line.


`r backto("rosetta")`

`r issues()`

# References
