---
title: "Independent Samples t-test"
author: Marcello Gallucci
topic: ttest
category: rosetta
nickname: ros_ttest_ind
bibliography: 
     - bib.bib
link-citations: true
        
---


```{r echo=FALSE,results='hide'}
library(mcdocs)
mcdocs_init()
```

# Validation

`r version("0.1.0")` 

Here we compare the results of `r modulename()` with other software that performs power analysis. In particular, we will compare our
results with [R pwr package](https://cran.r-project.org/web/packages/pwr/pwr.pdf) and [G*Power](https://www.psychologie.hhu.de/arbeitsgruppen/allgemeine-psychologie-und-arbeitspsychologie/gpower). 


# Example 1 : Power

:::{ .adm .adm-setup}
Setup

*   Aim = Power
*   Expected $\delta$ = .5
*   N per group = 30
*   Alpha = .05
*   Tails: Two

:::

## PAMLj

`r pic("rosetta/pics/ttest/ind.1.png")`

The expected power is .4779.

## R

```{r }

pwr::pwr.t.test(n=30,d=.5,sig.level=.05)

```

## G*Power

`r pic("rosetta/pics/ttest/gpower.ind.1.png")`


Results are the same


# Example 2: Sample size

:::{ .adm .adm-setup}
Setup

*   Aim = N
*   Expected $\delta$ = .4
*   Power = .90
*   Alpha = .05
*   Tails = Two
:::    

## PAMLj

`r pic("rosetta/pics/ttest/ind.2.png")`

The expected N is 266, 113 per group.
## R

```{r }

pwr::pwr.t.test(d=.4,power=.9,sig.level=.05)


```

## G*Power

`r pic("rosetta/pics/ttest/gpower.ind.2.png")`

Results are the same.

# Example 3: Effect size

:::{ .adm .adm-setup}
Setup

*   Aim = minimal effect size
*   power = .95
*   N per group = 52
*   Alpha = .01

:::

## PAMLj

`r pic("rosetta/pics/ttest/ind.3.png")`

## R


```{r }

pwr::pwr.t.test(n=52,power=.95,sig.level=.01)

```


## G*Power

`r pic("rosetta/pics/ttest/gpower.ind.3.png")`

Again, rounding a the third decimal place, results are the same.

# Example 4: One-tail

:::{ .adm .adm-setup}
Setup

*   Aim = N
*   Expected $\delta$ = .4
*   Power = .90
*   Alpha = .05
*   Tails = One

:::
### PAMLj

`r pic("rosetta/pics/ttest/ind.4.png")`


### R

```{r }

pwr::pwr.t.test(d=.4,power=.90,sig.level=.05,alternative="greater")

```

### G*power

`r pic("rosetta/pics/ttest/gpower.ind.4.png")`

`r backto("rosetta","ttest")` 


`r issues()`



# References