---
title: "General Linear Model"
topic: samplesize
category: rosetta
nickname: ros_glm_n
output: 
  html_document:
     includes:
         in_header: ganalytics.txt
     toc: true
     toc_float:
        collapsed: false
bibliography: 
     - bib.bib
link-citations: true
        
---


```{r echo=FALSE,results='hide'}
library(mcdocs)
mcdocs_init()
```

# Power Analysis: Minimum Sample Size 

`r version("0.1.0")` 

Here we compare the results of `r modulename()` with other software that performs power analysis. In particular, we will compare our results with [R pwr package](https://cran.r-project.org/web/packages/pwr/pwr.pdf) and [G*Power](https://www.psychologie.hhu.de/arbeitsgruppen/allgemeine-psychologie-und-arbeitspsychologie/gpower). 

We use a few examples from [G*Power manual](https://www.psychologie.hhu.de/arbeitsgruppen/allgemeine-psychologie-und-arbeitspsychologie/gpower), then we test some other design and analysis. In G*Power manual the aim of the analysis is to find the post hoc power, here we look for the minimum sample size given the power obtained in the G*Power manual. So, this is not a tutorial on how to conduct a power analysis (the required power is sometimes a non-sense), is a testing exercise to evaluate the correctness of the results.

# Multiple regression

<i> Multiple Regression - omnibus (deviation of R 2 from zero), fixed model, p. 33 </i>

Here we estimate the post-hoc power for the F-test associated to the $R^2$ of a model with 5 predictors. Five predictors means that the model degrees of freedom are 5. Because the test is on the whole model $R^2$, also the the effect DF are 5. The $R^2$ is transformed into the $f^2$ effect size with the simple formula $f^2=R^2/(1-R^2)$.

*   Aim = Sample Size
*   Expected $R^2$    = .10
*   $f^2$              = `r round(0.10/(1-.10),digits=4)`
*   power           = .6735
*   N = 95
*   Effect DF       = 5
*   Model  DF       = 5
*   Alpha           = .05

### G*Power

`r pic("rosetta/pics/gpower.n.glm.1.png")`

    
### R

In R, we can use the `pwr.f2.test` command, just plugging in the effect size $f^2$ and the effect DF $u=5$. However, $N$ should be inserted as the residuals degrees of freedom, thus we have $v=N-u-1=95-5-1=89$

```{r }

(res<-pwr::pwr.f2.test(f2=.1111,u=5, power=.6735,sig.level=.05))

N<-round((res$v+res$u+1))
N
```

### PAMLj

In `r modulename()` we can use both `r opt("Partial Eta-Squared")` or `r opt("Eta-Squared")`, because in the case of the $R^2$ the two effect size are identical. For simplicity, we use `r opt("Partial Eta-Squared")`

`r pic("rosetta/pics/glm.n.1.png")`

Had we used `r opt("Eta-Squared")`, we simply put the $R^2$ to test both in the effect size and the $R^2$ field, obtaining the same results.

`r pic("rosetta/pics/glm.n.1.2.png")`


# Increase in $R^2$

<i> F test: Multiple Regression - special (increase of R2), fixed model , p. 37 </i>

The first example examines a linear model (regression) with $R_f^2=.30$ with 9 predictors and compare it to a nested model with $R_n^2=.25$ with 5 predictors. Basically, the test is aimed at testing the variance explained by 4 predictors present in the full model but not the nested model. The variance explained by the 4 predictors is therefore $R_f^2-R_n^2=.30-.25=.05$.  This is also called $\Delta R^2$, but more formally is a $\eta^2$ with 4 degrees of freedom evaluated in a model with 9 degrees of freedom.

The $\eta^2$ is transformed into the $f^2$ effect size with the simple formula $f^2=(R_f^2-R_n^2)/(1-R_f^2)$.

*   Aim = Sample Size
*   $R_f^2$    = .30
*   $R_n^2$    = .25
*   $\eta^2$    =  0.05
*   $f^2$              = `r round(0.05/(1-.30),digits=5)`
*   Power = .241296
*   Effect DF       = 4
*   Model  DF       = 9
*   Alpha           = .01

### G*Power


`r pic("rosetta/pics/gpower.n.glm.2.png")`

### R

In R, we can use the `pwr.f2.test` command, just plugging in the effect size $f^2$ and the effect DF $u=4$. The final $N$ will results as $N=v+u+1$

```{r }
f2=.05/(1-.30)
u<-4
m<-9
power<- .241296
(res<-pwr::pwr.f2.test(f2=f2,u=u, power=power,sig.level=.01))
N<-round((res$v+u+1))
N

```


### PAMLj

Results are perfectly in line with other software.

`r pic("rosetta/pics/glm.n.2.png")`


# ANOVA 

<i> F test: Fixed effects ANOVA - special, main effects and interaction p. 27</i>

Here we have an ANOVA design $A \times B \times C$ with $A$ and $B$ having 3 levels and $C$ with 4 levels. Thus the overall model features 36 groups, with $A$ and $B$ effects having 2 degrees of freedom, and $C$ having 3 DF. The researcher is interested in the sample size required to achieve a power of $.475627$ of the test associated with the interaction $A*B$, with a partial Eta-squared $p\eta^2=0.05665751$. The interaction will have $DF=2*2=4$ degrees of freedom, and the whole model will have $DF=35$, the number of groups minus one. Here are the parameters. The resulting sample size should be $N = 90$.


*   Aim = Sample Size
*   $p\eta^2$   = 0.05665751
*   $f$              = `r round(sqrt(0.05665751/(1-0.05665751)),digits=5)`
*   Effect DF       = 4
*   Model  DF       = 35
*   Power           =  .475627
*   Alpha           = .01

### G*Power

First notice that for this problem, G*Power offers  _F test: Fixed effects ANOVA_ routine, which requires an $f$ as the effect size . $f$ is simply the square root of $f^2$ so it can be computed from the partial Eta-squared as  $f=\sqrt{p\eta^2/(1-p\eta^2)}$ . Based on the G*Power example, the resulting sample size should be N=108.

Plugging in the parameters we obtain:

`r pic("rosetta/pics/gpower.n.glm.3.png")`


### R

As we have seen before, `pwr.f2.test` command will underestimate the power due to the fact that the non centrality parameter is smaller than the one used in G*Power.  Nonetheless, we can get an estimation of the power increasing the effect size $f^2$ such as to match the correct NCP is $\lambda=Nf^2$. The effect DF $u=4$ and the model DF are $m=35$, yielding $v=N-m-1=108-35-1=72$

```{r }
f<-0.2450722
u<-4
m<-35
power=.475627
#N=108
#v<-N-m-1
#f2<-f^2*N/(u+v+1)
(res<-pwr::pwr.f2.test(f2=f2,u=4, power=power,sig.level=.05))
N<-round(res$v+m+1)
N
```

We obtained the same results as in G*Power

### PAMLj

We plug in the partial Eta-squared and the required parameters, and we obtain the same results as before.

`r pic("rosetta/pics/glm.n.3.png")`


`r issues()`

# Back to `r link_pages(nickname="rosetta")`


# References
'