---
title: "Correlation"
topic: correlation
category: rosetta
nickname: ros_correlation
output: 
  html_document:
     includes:
         in_header: ganalytics.txt
     toc: true
     toc_float:
        collapsed: false
bibliography: 
     - bib.bib
link-citations: true
        
---


```{r echo=FALSE,results='hide'}
library(mcdocs)
mcdocs_init()
```

# Power Analysis

`r version("0.1.0")` 

Here we compare the results of `r modulename()` with other software that performs power analysis. In particular, we will compare our
results with [R pwr package](https://cran.r-project.org/web/packages/pwr/pwr.pdf) and [G*Power](https://www.psychologie.hhu.de/arbeitsgruppen/allgemeine-psychologie-und-arbeitspsychologie/gpower). 


# Correlation

For Pearson and Spearman correlation, we can compare `r modulename()` with `R pwr package`. Consider, however, that for this task `r modulename()` employs `R pwr package` under the hood, so the results are obviously in line. Both packages use the  [inverse hyperbolic tangent](https://en.wikipedia.org/wiki/Inverse_hyperbolic_functions#atanh) transformation [@cohen1988spa].

## Example 1

*   Aim = Sample size 
*   Expected r = .3
*   Required power = .8
*   Alpha = .05

    
### R

```{r }

pwr::pwr.r.test(r=.3,power=.8,sig.level=.05)

```

### PAMLj

`r pic("rosetta/pics/r.1.png")`

If we round the `pwr` package results, they are the same

## Example 2

*   Aim = power
*   Expected r = .4
*   N = 52
*   Alpha = .05

    
### R


```{r }

pwr::pwr.r.test(r=.4,n=52,sig.level=.05)

```

### PAMLj

`r pic("rosetta/pics/r.2.png")`

Results are the same.

## Example 3

*   Aim = minimal effect size
*   power = .95
*   N = 52
*   Alpha = .01

### R


```{r }

pwr::pwr.r.test(n=52,power=.95,sig.level=.01)

```

### PAMLj

`r pic("rosetta/pics/r.3.png")`

Results are the same.

## Example 4

*   Aim = Sample size
*   power = .95
*   Expected r = .45
*   Alpha = .01
*   Tails = "two.sided"

### R


```{r }

pwr::pwr.r.test(r=.45,power=.95,sig.level=.01,alternative="greater")

```

### PAMLj

`r pic("rosetta/pics/r.4.png")`

Results are the same.

# Navigate

Back to `r link_pages(nickname="rosetta")`


`r issues()`

# References
'