---
title: "Independent Samples Proportions"
author: Marcello Gallucci
topic: prop
category: rosetta
nickname: ros_prop_ind
bibliography: 
     - bib.bib
link-citations: true
        
---


```{r echo=FALSE,results='hide'}
library(mcdocs)
mcdocs_init()
```

# Validation

`r version("0.2.0")` 

Here we compare the results of `r modulename()` with other software that performs power analysis. In particular, we will compare our
results with [R pwr package](https://cran.r-project.org/web/packages/pwr/pwr.pdf) and [G*Power](https://www.psychologie.hhu.de/arbeitsgruppen/allgemeine-psychologie-und-arbeitspsychologie/gpower). 

Each software employs a z-test after arch transforming the proportions.

# Example 1: Sample size

:::{ .adm .adm-setup}
Setup

*   Aim   = N
*   P1    = .6
*   P2    = .5
*   Power = .90
*   Alpha = .05
*   Tails = Two
:::    

## PAMLj

`r pic("rosetta/pics/prop/ind.1.png")`

The expected N is 1038, 519 per group.

## R

```{r }

(h<-pwr::ES.h(.6,.5))
pwr::pwr.2p.test(h=h,power=.9,sig.level=.05)

```
`ceiling` the results we obtain 519 cases per group.

## G*Power

`r pic("rosetta/pics/prop/gpower.ind.1.png")`

Results are the same.

# Example 2 : Power

:::{ .adm .adm-setup}
Setup

*   Aim   = Power
*   P1    = .6
*   P2    = .5
*   N per group = 200
*   Alpha = .05
*   Tails = Two

:::

## PAMLj

`r pic("rosetta/pics/prop/ind.2.png")`


## R

```{r }

(h<-pwr::ES.h(.6,.5))
pwr::pwr.2p.test(n=200,h=h,sig.level=.05)

```

## G*Power

`r pic("rosetta/pics/prop/gpower.ind.2.png")`


Results are almost the same, with a trivial difference of `r 0.5214145-.5301179`


# Example 3: Unbalanced designs

:::{ .adm .adm-setup}
Setup

*   Aim = minimal effect size
*   power = .95
*   N per group = 52
*   Alpha = .01

:::

## PAMLj

`r pic("rosetta/pics/prop/ind.3.png")`

## R


```{r }

(h<-pwr::ES.h(.6,.5))
pwr::pwr.2p2n.test(n1=50,n2=60,h=h,sig.level=.05)

```


## G*Power

`r pic("rosetta/pics/prop/gpower.ind.3.png")`

Again,  results are the very similar.


`r backto("rosetta","prop")` 


`r issues()`



# References