---
title: "Independent sample T-test: Equivalence tests"
topic: ttest
category: rosetta
nickname: ttest_ind_equi
output: 
  html_document:
     includes:
         in_header: ganalytics.txt
     toc: true
     toc_float:
        collapsed: false
bibliography: 
     - bib.bib
link-citations: true
        
---


```{r echo=FALSE,results='hide'}
library(mcdocs)
mcdocs_init()

```

# Power Analysis

`r version("0.2.0")` 

Here we compare the results of `r modulename()` with published results and other software that performs equivalence tests for the t-test.

# Independent sample t-test

For the independent t-test, we will refer to @lakens2017equivalence `r ext_url("Table 1","https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5502906/table/table1-1948550617697177/?report=objectonly")`
, in which sample size are reported for an equivalence test with expected value $\delta=0$ and varying "Bound", in `r modulename()` refereed to as 'equivalence limit'. @lakens2017equivalence Table 1 reports results for $\alpha=.05$ and $\alpha=.01$.  The table (available `r ext_url("here","https://www.ncbi.nlm.nih.gov/pmc/articles/PMC5502906/table/table1-1948550617697177/?report=objectonly")`) is reproduced below.

`r pic("rosetta/pics/ttest_ind_equi_1.png")`

The table reports the required N to obtain a sufficient power (80% or 90%) given a certain $\alpha$ and a certain limit $\Delta$ of equivalence.


### PAMLj

For the firs test, we insert an `r opt("Equivalence limit")` of $.1$, an expected $\delta=0$, power=.80 and $\alpha=.05$.
`r pic("rosetta/pics/ttest_ind_equi_2.png")`
`r pic("rosetta/pics/ttest_ind_equi_3.png")`

### R

```{r }

pwr::pwr.r.test(r=.3,power=.8,sig.level=.05)

```

### G*Power

`r pic("rosetta/pics/gpower.r.1.png")`


If we round the results, they are the same


## Example 2

*   Aim = power
*   Expected r = .4
*   N = 52
*   Alpha = .05

    

### PAMLj

`r pic("rosetta/pics/r.2.png")`

### R

```{r }

pwr::pwr.r.test(r=.4,n=52,sig.level=.05)

```

### G*Power

`r pic("rosetta/pics/gpower.r.2.png")`

Results are the same at the third decimal place, which can be consider quite good.

## Example 3

*   Aim = minimal effect size
*   power = .95
*   N = 52
*   Alpha = .01

### PAMLj

`r pic("rosetta/pics/r.3.png")`

### R


```{r }

pwr::pwr.r.test(n=52,power=.95,sig.level=.01)

```


### G*Power

`r pic("rosetta/pics/gpower.r.3.png")`

Again, rounding a the third decimal place, results are the same.

## Example 4

*   Aim = Sample size
*   power = .95
*   Expected r = .45
*   Alpha = .01
*   Tails = "two.sided"

### PAMLj

`r pic("rosetta/pics/r.4.png")`


### R

```{r }

pwr::pwr.r.test(r=.45,power=.95,sig.level=.01,alternative="greater")

```

### G*power

`r pic("rosetta/pics/gpower.r.4.png")`

In this case, notice that `r modulename()` yield 69, which is rounded for 69.45, whereas G*Power round it up to 70. Thus, results are quite in line.

# Navigate

Back to `r link_pages(nickname="rosetta")`


`r issues()`

# References
'