---
title: "SEM power analysis"
author: "Marcello Gallucci"
topic: sem
output: pdf_document
bibliography: bib.bib
link-citations: yes
editor_options:
  chunk_output_type: console
category: help
---


```{r echo=FALSE,results='hide'}
library(mcdocs)
mcdocs_init()

```

`r version("0.6.0")`

`r pic("pics/sem/ui.png")`


`SEM power analysis` enables the computation of power parameters (such as required N and power) for a wide range of structural equation models. It can be used to estimate power parameters based on both chi-square constraints and `Monte Carlo` methods.

The module allows users to specify the prospective model using the syntax of the [lavaan R package](https://lavaan.ugent.be/), with some modifications necessary to input the expected parameters. The results are partially obtained using the [semPower R package](https://CRAN.R-project.org/package=semPower), though several solutions are also implemented natively within the module.

:::{ .adm .adm-note}
Important

The capabilities (so far) of the module is to obtain power parameters associated with tests of single coefficients, or combinations of coefficients.

:::

# Input

The crucial aspect of the module usage is to input the expected model. One can start from a lavaan syntax model, such as 

```{r eval=FALSE}

# latent variables 
   ind60 =~ x1 + x2 + x3 
   dem60 =~ y1 + y2 + y3 + y4 
   dem65 =~ y5 + y6 + y7 + y8 
 # regressions
   dem60 ~ ind60 
   dem65 ~ ind60 + dem60 
  # residual correlations
   y1 ~~ y5    

```


To input the expected coefficients, one multiplies each variable for the associated coefficient

```{r eval=FALSE}

# latent variables 
    ind60 =~ .6*x1 + .6*x2 + .6*x3
    dem60 =~ .8*y1 + .8*y2 + .8*y3 + .8*y4
    dem65 =~ .8*y5 + .8*y6 + .8*y7 + .8*y8
  # regressions
    dem60 ~ .3*a*ind60
    dem65 ~  .4*b*ind60 + .2*c*dem60
  # residual correlations
    y1 ~~ .1*y5    

```


In this example, we specify that the latent variable `ind60` should have loadings equal to $0.6$, and both the latent variables `dem60` and `dem65` should have loadings equal to $0.9$. Additionally, the regression coefficients for the latent variables are set to $0.3$, $0.4$, and $0.2$, respectively. The scale of the coefficients should be chosen under the assumption that all observed variables are standardized.

Another requirement is to specify labels for the parameters we wish to test. In this example, we have labeled the three latent variables as $a$, $b$, and $c$, respectively. This allows us to impose constraints on them and compute the required sample size (or power) to obtain a significant test (given a specified $\alpha$) for those particular coefficients or combinations of coefficients. This is achieved by constraining the coefficients to zero.

```{r eval=FALSE}

# latent variables 
    ind60 =~ .6*x1 + .6*x2 + .6*x3
    dem60 =~ .8*y1 + .8*y2 + .8*y3 + .8*y4
    dem65 =~ .8*y5 + .8*y6 + .8*y7 + .8*y8
  # regressions
    dem60 ~ .3*a*ind60
    dem65 ~  .4*b*ind60 + .2*c*dem60
  # residual correlations
    y1 ~~ .1*y5
  # tests
    a == 0
    
```


In this case, we are requesting the power parameter associated with the coefficient $a*ind60 = 0.3$. The module will compute the required parameters to obtain a statistically significant test by comparing two models: one where the coefficient of `ind60` is freely estimated and another where the coefficient of `ind60` is constrained to $0$. In the module, this model looks like this:

`r pic("pics/sem/input.1.1.png")`

After checking the usual input values (Type I error, required power, etc)

`r pic("pics/sem/input.1.2.png")`

# Results

`r pic("pics/sem/input.1.1.png")`


The main table reports the required N to achieve a statistically significant result when testing the `ind60` coefficient. Additionally, we are provided with the path diagram to verify whether the model aligns with our expectations. Admittedly, the path diagram may appear a bit cluttered at times, but its purpose is solely to confirm the model's structure, not to look aesthetically pleasing.


`r pic("pics/sem/input.1.2.png")`

# Options

## `r opt_panel("Main")`

`r opt_label("Parameters")`

`r panel_options('pamlsem','power_parameters')`

`r opt_label("Model Method & Options")`

`r panel_options('pamlsem','model_options')`

## `r opt_panel("Sensitivity Analysis")`

Sensitivity analysis, which explores different plots of possible parameter combinations, can be conducted as in any other `r modulename()` sub-module. Please visit the `r link_pages(nickname="sensitivity")` page for more details. For the SEM sub-module, sensitivity analysis is available only for varying N values. When the Monte Carlo method is selected, sensitivity analysis is performed using the analytic method; otherwise, the computation time would be extremely long.


## `r opt_panel("Power Parameters plot")`

```{r child_sensitivity_link, child='commons/powerplot_chunk.Rmd'}
```



## `r opt_panel("Options")`

`r panel_options('pamlsem','panel_options',labels=T)`



`r backto()`

`r issues()`
