---
title: "Paired samples t-test"
author: Marcello Gallucci 
topic: ttest
category: subhelp
order: 2
bibliography: 
     - bib.bib
link-citations: true
        
---


```{r echo=FALSE,results='hide'}
library(mcdocs)
mcdocs_init()
```

# Aim

`r panel_options('pamlttest','aim')`

# Input

`r panel_options('pamlttest','mode', labels=T,mode="ttestpaired")`

`r panel_options('pamlttest','box_parameters', labels=T)`

`r panel_options('pamlttest','box_equi_test', labels=T)`


```{r child_sensitivity_link, child='commons/sensitivity_chunk.Rmd'}
```

# Effect Size Conversion

With this panel is possible to estimate the expected effect size based on observed t-test, taken from empirical studies or other empirical sources.


`r panel_options('pamlttest','convert_es_ttestpaired', labels=T)`


It should be noted that `r modulename()` uses **Cohen’s** $d_z$ as the effect size for power analysis. This is defined as the **mean of the difference scores divided by their standard deviation** [@cohen1988spa]. Some authors instead suggest starting from **Cohen’s** $d$ for independent samples, defined as the **mean difference divided by the square root of the sum of the variances** of the two measures. If you want to convert this independent-samples $d$ into a paired-samples $d_z$, accounting for the expected correlation $r$ between measures, you can use the `r opt_label("Convert to dz")`, which applies the following formula:

$$
d_z = \frac{d}{\sqrt{2(1 - r)}}
$$


`r version("0.7.0")`


`r issues()`

`r backto("ttest")`